---
title: "RProjectRedo"
output: github_document
date: "2025-01-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(ISwR)
library(car)
library(ISLR)
options(scipen=999)
```

The variables in this dataset are:
education: Average years of education
income: Average inome
women: Percent of women
prestige: Pineo-Porter prestige score
census: Canada Census occupational code
type: The type of job. Blue collar, white collar, and professional are the types.

```{r}
prestige = read.csv("C:/Users/katek/OneDrive - University of South Florida/Desktop/NCF/Fall 2024/Linear models/Prestige.txt", sep="")
prestige
```

Removing the "census" column, as it's just an ID number. Also removing the "NA" values.
```{r}
prestige=na.omit(prestige)
prestigeclean=prestige[,-5]
```

Creating and interpreting a linear model:
```{r}
prmodel=lm(prestige~education+income+women+type,data=prestigeclean)
summary(prmodel)
```
The equation for this model: prestige = -.81 + 3.66*education + .001*income + .0064*women + 5.91*typeprof - 2.92*typewc
Education and income have a significant effect of prestige, since their P-values are less than .05. For every year of education, the career's prestige increases by 3.66 on average, holding all other variables constant. For every dollar of income, the career's prestige increases by .001 on average, holding all other variables constant.

Model with all variable interactions:
```{r}
prmodelint=lm(prestige~.^2,data=prestigeclean)
summary(prmodelint)
```
Income:typeprof is a significant interaction, since its p-value is less than .05.

Creating model with the interaction term:
```{r}
printmodel=lm(prestige~education+income+women+type+income:type,data=prestigeclean)
```

Plotting the interaction:
```{r}
ggplot(printmodel, aes(x=income, y=prestige, color=type)) + geom_smooth(method='lm')
```
The slope for the relationship between income and prestige significantly changes when the measurements are taken from blue collar jobs.

```{r}
summary(printmodel)
```

Performing an anova test between the model with and without the interaction term and interpreting the result:
```{r}
anova(prmodel, printmodel)
```
The model with the interaction term has more significance than the one without.

Finding outliers:
```{r}
plot(cooks.distance(printmodel))
text(cooks.distance(printmodel))
```

Removing outliers:
```{r}
prestigeout=prestigeclean[-c(2,31,50,78),]
```

Creating and interpreting the linear model:
```{r}
proutmodel=lm(prestige~education+income+women+type+income:type,data=prestigeout)
summary(proutmodel)
```
The equation for this model: prestige = -5.82 + 2.19*education + .004*income + .057*women + 31.12*typeprof + 5.64*typewc - .003*income:typeprof - .0011*income:typewc
Education, income, typeprof, and income:typeprof significantly effect prestige.
For every year of education, the career's prestige increases by 2.19 on average, holding all other variables constant. For every dollar of income, the prestige increases by .004 for blue collar jobs, .0029 for white collar jobs, and .001 for professional jobs.
Additionally, the intercepts are -5.82, -.18, and 25.3 for blue collar, white collar, and professional jobs, respectively.
Equation for professional jobs: (-5.82+31.12*typeprof) + 2.19*education + (.004*income-.003*income:typeprof) + .057*women = 25.3 + 2.19*education + .001*income + .057*women
Equation for white collar jobs: (-5.82+5.64*typewc) + 2.19*education + (.004income-.0011*income:typewc) + .057*women = -.18 + 2.19*education + .0029*income + .057*women
Equation for blue collar jobs: -5.82 + 2.19*education + .004*income + .057*women

Let's compare models. For reference, the model with the interaction term is on top and the model on the bottom has the interaction term and no outliers.

```{r}
summary(printmodel)$fstatistic[1]
summary(proutmodel)$fstatistic[1]
```
The F-statistic is much higher with the new model, so it's even more likely at least 1 variable affects prestige.

```{r}
summary(printmodel)$r.squared
summary(proutmodel)$r.squared
```
For the model with the outliers, the R-squared was 87.46%, meaning of the variance in prestige was explained by the other variables. Without them,  89.82% of the variance is explained by the other variables, making it a better model in that regard.

```{r}
sigma(printmodel)
sigma(proutmodel)
```
The original model had an RSE (residual standars error) of 6.28, meaning the model's estimate of the prestige score was off by 6.28 on average. The model with the interaction term was only off by 5.69, making it a better fit.

Finding and interpreting confidence intervals:
```{r}
confint(proutmodel)[1,] 
```
I am 95% confident that for careers with 0 years of education, 0% women, and $0 income, the prestige will be between 0 and 3.57.
For this interpretation, I cut it off at 0, because a negative prestige rating doesn't exist.

```{r}
confint(proutmodel)[2,]
confint(proutmodel)[3,]
confint(proutmodel)[4,]
```
I am 95% confident that for every year of education, the prestige increases by between 1.04 and 3.35 points on average.
I am 95% confident that for every dollar of income, the prestige increases by between .0029 and .0051 points on average.
I am 95% confident that the slope of the relationship between the percent of women and prestige is between -.0019 and .12 points on average.

Scaling the numerical variables:
```{r}
prestigescaled=data.frame(scale(prestigeout[,1:4]))
prestigescaled$type=prestigeout$type
```
With the model using the scaled data, I can see what variable affects prestige the most.

```{r}
prscmodel=lm(prestige~education+income+women+type+income:type,data=prestigescaled)
summary(prscmodel)
```

The equation for this model: prestige = -.021 + .36*education + .89*income + .1*women + .62*typeprof - .096*typewc - .67*income:typeprof - .24*income:typewc
Income has the strongest effect on prestige, because the absolute value of its estimate is the highest. Listing the variables from strongest significant effect to weakest significant effect on prestige gives us: income, income:typeprof, typeprof, then education. Surprisingly, education doesn't matter as much as the type of job or income when it comes to the prestige score.

Equation for professional jobs: .6 + .36*education + .22*income + .1*women
Equation for white collar jobs: -.12 + .36*education + .65*income + .1*women
Equation for blue collar jobs: -.021 + .36*education + .89*income + .1*women
I will define base prestige as the estimated prestige score when education, women, and income are 0. Blue collar jobs refer to jobs that involve manual labor, white collar jobs are office jobs, and professional jobs require specialized training. When writing the equations for each career type, it showed that blue collar jobs have the lowest base prestige while professional jobs have the highest. This implies that a professional job is seen as more prestigious even if all variables are the same as a blue collar job. Despite this, if all variables between a blue collar and professional job are the same, with enough income, the blue collar job will be more prestigous. This makes sense, as the variable that affects a blue collar job's prestige the most is income. The variable that affects a professional job's prestige the most is education, which is likely relted to the fact it requires specialized training.
